---
title: "Stheatmap Vignette"
author: "Radhika Kulkarni"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stheatmap Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(stheatmap)
```


## Introduction

Stheatmap is a package that specifically creates visualizations for two sources of spatiotemporal data. The first source, which is data on the prevalence of Drought, is used in the package to visualize Percentage of County Area experiencing a certain level of severity of drought in a particular Geographic Region using a heatmap. The second source, data on the Oceanic Niño (ONI) values of the El Niño Southern Oscillation (ENSO) effect, is utilized by the package to create two plots: one a general smoothed trend curve, and the other a detailed barplot of the ENSO events and corresponding severities. 

The author and maintainer of stheatmap R package is Radhika Kulkarni. This
vignette describes the usage of stheatmap in R.


### Motivation for `stheatmap`

When we have a large volume of data associated with geographic locations, taken over a period of time, it is often difficult to see the relevant patterns inherent in the data.

In this package demo we visualize California Drought data taken from US Drought Monitor using a function that creates heatmaps generated using agglomerative hierarchical clustering with euclidean distance metric. We also visualize El Niño Southern Oscillation Data (ENSO) data using a rolling average smoother which captures the cyclical nature of ENSO (seasonal trends). These bigger picture visualizations can show us how the global and local variables interrelate with each other on the grand scale and can reveal some interesting associations in specific geographical areas or time points.


### Functions in the Package 

#### Two Functions, Three Plots

1. **Drought Function**: 
  - `drought_main(drought_data, start_date, end_date)`
  - **Plot**: Heatmap for Drought data
  
2. **ENSO Function**: 
  - `enso_main(data, StartDate, EndDate)`
  
  - **Plot 1** : Time Series Line Plot for Overall Trend at a glance
  
  - **Plot 2**: ENSO barplot using Rolling Avg Smoother

### Installation 

To begin, install and load `stheatmap`:

```{r, warning=FALSE}
library(stheatmap)
```



### Data Download

##### Drought Data

To generate the heatmap for Drought data, the user must obtain the data from the **US Drought Monitor** at [US Drought Monitor - Comprehensive Statistics](https://droughtmonitor.unl.edu/DmData/DataDownload/ComprehensiveStatistics.aspx).

#### Input Requirements for the `drought_main()` Function

1. **Download the Data**:
   - **Source**: US Drought Monitor
   - Specify the following parameters:
     - **Start and End Date**: Chosen by the user.
     - **Area Type**: State.
     - **Statistics Category**: Percent Area (percentage of the area of that county with the specified drought severity category).
     - **Statistics Type**: Cumulative (percentage of area adds up to 100%).
     - **Output Format**: CSV.

2. **Clean the Data**:
   - Retain only the following columns:
     - `MapDate`, `None`, `D0`,`D1`,`D2`,`D3`,`D4`,`ValidStart`.

```{r}
# Example code for cleaning data
clean_drought_data <- function(drought_data) {
  # Retain only the relevant columns
  drought_data <- drought_data %>%
    select(MapDate, None, D0, D1, D2, D3, D4, ValidStart)
  
  return(drought_data)
}
```


##### ENSO Data

To generate the barplot for ENSO data, the user must obtain the data table from the **National Weather Service, Climate Prediction Center** at [Cold & Warm Episodes by Season](https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/ONI_v5.php).This table can be converted to .csv or pasted in an Excel file to be used. 

**Note: The Data must retain the structure of the table on the website to be used as input to the package. If some data is unavailable (i.e. if current month is December, the rest of the months in the row for that Season will not be available yet), set the values in the csv/Excel file to NA.**



### Clustering Algorithm For Assigning Drought Severity

Each county-week observation is represented as a \(p\)-dimensional vector:
\[
x_i = (x_{i1}, x_{i2}, \dots, x_{ip})
\]
where \(x_{il}\) is the average percentage for drought severity category \(l\) in observation \(i\).

To measure dissimilarity between observations \(x_i\) and \(x_j\), we use the Euclidean distance:
\[
d(x_i, x_j) = \sqrt{\sum_{l=1}^p (x_{il} - x_{jl})^2}.
\]

For hierarchical clustering with average linkage, the distance between two clusters \(C_u\) and \(C_v\) is defined as the average of all pairwise distances between their members:
\[
D_{\text{avg}}(C_u, C_v) = \frac{1}{|C_u| \cdot |C_v|} \sum_{x_i \in C_u} \sum_{x_j \in C_v} d(x_i, x_j),
\]
where \(|C_u|\) and \(|C_v|\) are the sizes of clusters \(C_u\) and \(C_v\).

The algorithm proceeds by:

1. Initializing each observation as its own cluster.

2. Merging the two closest clusters based on the above distance metric.

3. Recomputing distances until \(K=6\) clusters remain.

Cutting the resulting dendrogram at the height corresponding to \(K\) clusters provides the final grouping, and each cluster is assigned a representative drought category reflecting its predominant severity level composition.




### Demo and Introduction to the Data  

The Drought data is subset to be spatially in California, and the El Niño/La Niña oscillation cycle values are a global weather pattern. To investigate how these global and local weather patterns are associated, time series will be over the time period of 20 years, from January 1, 2001 - June 1, 2021. If wishing to visually compare another state's drought patterns with the ENSO effect, you may download the data for that specific state.


Both the Drought data and the ENSO data in this demo are available as in the Data folder as .Rda files, and they can be called through objects Drought and ENSO, respectively, in the R package. To use the datasets, simply load the datasets using the following commands:


```{r loaddata}
load(system.file("data/ENSO.Rda", package = "stheatmap"))
load(system.file("data/Drought.Rda", package = "stheatmap"))

```



#### California Drought Data, Jan 2001 - Jun 2021 (1066 weeks):

Let us have a look at the desired data structure to function `drought_main()` in the package: 

```{r headdrought}
# View the first few entries of the Drought dataset
head(Drought)

#get number of rows and columns
dim(Drought)
```

There are `r dim(Drought)[1]` records of drought in California in the time range of 1066 weeks, and 58 counties (58 x 1066). 


#### El Niño Southern Oscillation Effect Data, Jan 2001 - Jun 2021 (245 months):

Let us have a look at the desired data structure to function `enso_main()` in the package: 

```{r headenso}
# View the first few entries of the ENSO dataset
head(ENSO)

# View the last few entries of the ENSO dataset, including NAs
tail(ENSO)

#get number of rows and columns
dim(ENSO)
```

There are `r dim(ENSO)[1]` Seasons from 1950-1951 until 2021-2022. 

##### Plotting the Heatmap for the Drought Data:

Below we plot the heatmap for drought data taken over the range of Jan 2001 - June 2021. 
The x-axis are the names of the 58 counties in California, and the y-axis are the weeks over which the data was taken, from 1 to 1066. The color bar to the right denotes the severity of the drought; darkest red means most severe drought (category D4), while darkest green color means no drought (category None). 

```{r plotheatmap, fig.width=8, fig.height=6, out.width='80%', out.height='80%'}
# Plot the heatmap for drought
out <- drought_main(Drought, start_date = "2001-01-01", end_date = "2006-06-01")


```


From the plot above we can observe that Week 0 corresponds to January 1, 2001, and Week 1065 corresponds to June 1, 2021, establishing a continuous temporal span of twenty years along the heatmap’s vertical axis. During the years in the beginning of the series (2001–2005), the visualization mostly exhibits green tiles, signifying "no drought" (None) conditions and indicating relatively mild climate patterns across most counties. As the timeline progresses into the mid-2010s, which is a period well-known to have been experiencing severe drought episodes in California, the color values shift toward yellows, oranges, and reds, reflecting an escalation to more severe drought levels (D2–D4). Counties situated in inland, agricultural regions (i.e. Central Valley) frequently manifest these higher-intensity drought color values for longer durations, suggesting a greater susceptibility to chronic moisture deficits. In contrast, coastal and northern counties tend to display intermittent or less intense drought signals (e.g., persistent greens and lighter yellows), suggesting that these areas may have benefited from comparatively milder drought conditions, perhaps due to their proximity to the ocean and less drought prone-climates.

##### Plotting the Time Series Plot for the ENSO Data:

Our first plot from `enso_main()` is a time series plot of the Oceanic Niño values. This plot gives the researcher the overall trend of the severity of ENSO events but not the type or named category of the severity. 

Here we plot the series for ENSO data taken over the range of Jan 2001 - June 2021. 
The x-axis are the years-months over which the data was taken, from 2001-01 to 2021-06, and the y-axis is the monthly oceanic Niño index value average (smoothed data using three month rolling average). 

```{r plotensotimeseries, fig.width=8, fig.height=8, out.width='80%', out.height='80%'}
# Plot the barplot for enso, first index is the desired plot
enso_main(ENSO,  "2001-01-01", "2021-06-01")[[1]]


```

Below the 0 line indicates a La Niña event (cold water), and above the 0 line is an El Niño event (warm water). Note that there is a general trend of increasing severity of events from 2001 to 2016 for both El Niño and La Niña. This ties into the climate change theory that over recent history, climate events and phenomenon exhibit more extreme behaviors. 



##### Plotting the Barplot for the ENSO Data:

Below we plot the barplot for ENSO data taken over the range of Jan 2001 - June 2021. 
The x-axis is the monthly Oceanic Niño index value average (smoothed data using three month rolling average), and the y-axis are the years-months over which the data was taken, from 2001-01 to 2021-06. The legend to the right denotes the severity of the ENSO event. Red color indicates warm water events in the Niño 3.4 region, and blue color indicates cold water events in the region. 


```{r plotensobar, fig.width=8, fig.height=20, out.width='80%', out.height='80%'}
# Plot the barplot for enso, second index is the desired plot
enso_main(ENSO,  "2001-01-01", "2021-06-01")[[2]]


```


From 2001 through roughly 2003, the plot shows mostly neutral conditions with brief periods of weak El Niño (light red) and neutral (gray) anomalies, indicating only minor departures from average sea surface temperatures in the Niño 3.4 region. In the years 2004–2007, the pattern continues to be mixed but generally subdued, with episodes shifting between weak El Niño (pale reds) and occasional weak La Niña (lighter blues), without any intensely strong events. By 2007–2008, however, a more pronounced La Niña emerges (darker blues), signaling notably cooler ocean temperatures and a stronger ENSO signal. The period from late 2009 into early 2010 reverses course with a moderate El Niño (medium reds), which then transitions again to a series of strong La Niña conditions spanning roughly 2010–2012. Following a relatively neutral phase in the early 2010s, a prominent peak appears in 2015–2016 with a very strong El Niño event (deep reds), one of the most significant warm anomalies in the twenty year period. After this peak, ENSO conditions fluctuate between neutral and weaker La Niña phases (2016–2021), with fewer sustained or extreme departures by the end of the series.

When comparing these ENSO fluctuations to California’s drought timeline, the connection is not immediate. For example, the intense 2015–2016 El Niño did not immediately end the state’s prolonged drought. Instead, there was a lag of several months before increased precipitation was observed. Similarly, cooler La Niña events generally are associated with drier conditions, but this often happens in the previous wet season, rather than aligning perfectly with the timing of the ocean temperature anomaly. So, while the ENSO phases set a broad climate context, with El Niño often being a predictor of wet winters and La Niña toward drier ones, they do not guarantee immediate changes in California’s drought status. Instead, the influence is visible after a lag effect, since local atmospheric patterns, precipitation cycles, and regional factors translate oceanic anomalies into tangible climatic effects on the ground.


